Parental Migration and Left-Behind Adolescent's School Performance

### 2022-06-03 MScThesisText.Rmd
add table display in Rmd

```{r setup, include = FALSE}
library("papaja")
r_refs("references.bib")
```

@@

Parental migration shapes child-and-adolescence development by exerting different effects that work oppositely. Consider children's human capital accumulation. Labor migrants send remittances back home, boosting the family income. Higher income can fund children's schooling as well as improve their nutritional and living conditions. All these may enhance children's educational outcome. The inflow of remittances, however, cannot substitute for the loss of parental care. Parental migration halts the day-to-day, face-to-face contact between parents and children. It decreases the amount and quality of parental guidance, protection, and support. All these may hinder children's school performance. It is uncertain how these two opposing effects affect children's educational performance, which merits empirical investigation.

A challenge arises in the methodology for assessing the causal relation between parental migration and children's education. It is unfeasible and unethical to design an experiment that randomly assigns parents to migrate without their children. Statistical advances in causal inference provides an alternative. We can infer causality from a quasi-experimental design -- that is, only if the cause is as good as randomly assigned to individuals in the analysis, conditional on identification assumptions.

@@

To what extent do parental migration affect the cognitive development and academic performance of left-behind children? This thesis aims to address the shortcomings in the existing literature. I analyze the data from a nationally representative two-wave panel survey. The sample includes children in both urban and rural areas. I develop a difference-in-differences design for causal inference. The dependent variables include children's cognitive test scores and their Chinese, mathematics, and English exam scores.

@@

Two competing theoretical frameworks, one of which is the resource generation model. According to this model, parental migration brings forth greater economic capital to improve household environment. Economic capital is, in Bourdieu's words, "immediately and directly convertible into money and may be institutionalized in the form of property rights" [-@bourdieu2002, p.16]. [@chang2019] provided evidence through interviews with migrant workers and their families. They said that labor migration can boost household income, as the city offers more jobs and higher pay than does the countryside.

The other theoretical model is the family disruption model. Children flourish under parental nurturing care. This term refers to the "stable environments that promote health and adequate nutrition, protect from threats, and provide opportunities for learning and responsive, emotionally supportive and developmentally enriching relationships" [@black2021, p.1]. Parental migration decreases nurturing care, thereby hindering children's developmental outcomes. The loss of nurturing care represents a decreased investment in children's human capital. Human capital is the stock of "skills, talents, health, and expertise" in people that enhance their productivity [@goldin2016, @botev2019]. In disadvantaged families, the migrant parent is likely the best educated adult in the household. Their absence makes it harder for children to overcome learning obstacles.

@@

I obtained the data from the China Education Panel Survey (CEPS). It is a nationally representative, school-based survey featuring junior high school students. The survey administrator is the National Survey Research Center at Renmin University of China. The survey's sampling adopted a multi-stage stratified probability proportional to size approach. Moreover, designed deliberately to focus on migrant children, the survey also oversampled those counties with large intranational migrant populations. The baseline survey took place in the 2013-2014 academic year with a sample size of about twenty thousand students in seventh and ninth grades. These students were nested in 438 classrooms of 112 schools in 28 urban districts or rural counties. I only kept the 9,449 observations of seventh grade students and their parents, who completed the follow-up survey in the 2014-2015 academic year.

@@

~~I explored two categories of adolescence development: academic and cognitive abilities. One key dependent variable, the student's cognitive skill, is measured by a fifteen-minute standardized test. The test covered three types of reasoning: verbal, visuospatial, and numerical. The raw scores of cognitive skill were then standardized. The other~~ set of dependent variables concern students' academic performance. This is measured by the mid-term exam scores in three subjects: Chinese, mathematics, and English. To facilitate comparison across schools and counties, I use standardized test scores instead of raw test scores.

The key independent variable, the parental migration status, was identified by two items in the parent survey. In both the baseline and the follow-up surveys, parents specified the family members living in the same household at the time. I identified seven types of migration arrangement: both parents absent, only father absent, only mother absent, father absent, mother absent, and both parents at home. I then removed the observations that reported other types of parental absence induced by divorce or death.

@@

I adopted `r cite_r("r-references.bib")` for all the analyses.

Following [@bai2018], I adopted a difference-in-differences design to analyze the educational outcome of children newly left behind by parents vis-a-vis that of children staying together with parents. The model is:

$$
\Delta score_{i,s} = \alpha + \beta \cdot migr_{i,s} + \gamma \cdot FE_{s} + \lambda \cdot score_{i,s;base} + \theta \cdot X_{i,s} + \varepsilon_{i,s}
$$

whereas for student i in school s, $\Delta score_{i,s}$ is the change in standardized test score between baseline and follow-up surveys, $migr_{i,s}$ is the treatment dummy variable, $FE_{s}$ is the school-level fixed effect, $score_{i,s;base}$ is the baseline exam score, and $X_{i,s}$ is the vector of covariates capturing the characteristics of children and their households. The control variables on children includes age, gender, ethnicity, self-rated health, hukou type, and only child. The control variables on household include the years of education of the best-educated parent, the number of books at home, and the internet assess at home.

@@

>>>>

```{r setup, include = FALSE}
load('.Rdata')
library("papaja")
r_refs("references.bib")
```

@@

Parental migration impacts child-and-adolescence development by exerting different effects that work oppositely [@vanhook2020]. Consider children's education. Labor migrants send remittances back home, boosting the family income. Higher income can fund children's schooling as well as improve their nutritional and living conditions. All these may enhance children's educational outcome. The inflow of remittances, however, cannot substitute for the loss of parental care. Parental migration halts the day-to-day, face-to-face contact between parents and children. It decreases the amount and quality of parental guidance, protection, and support. All these may hinder children's school performance. It is uncertain how these two opposing effects shape children's educational performance, which merits empirical investigation.

A challenge arises in the empirical assessment of the causal relation between parental migration and children's education. It is unfeasible and unethical to design an experiment that randomly assigns parents to migrate without their children. Statistical advances in causal inference provides an alternative. We can infer causality from a quasi-experimental design -- that is, only if the cause is as good as randomly assigned to individuals in the analysis, conditional on identification assumptions.

@@

To what extent do parental migration affect the cognitive development and academic performance of left-behind children in the short term? I analyze the data from a nationally representative two-wave panel survey. The sample includes children in both urban and rural areas. I develop a difference-in-differences design for causal inference. The dependent variables include children's cognitive test score and Chinese, mathematics, and English exam scores.

@@

Between parental migration and children's education, the mechanisms trace to two theoretical frameworks. One model, the generation model, holds that parental migration brings forth greater resources in the form of economic capital. Economic capital is, in Bourdieu's words, "immediately and directly convertible into money and may be institutionalized in the form of property rights" [-@bourdieu2002, p.16]. Economic capital is transferable to human capital, which is the stock of "skills, talents, health, and expertise" in people that enhance their productivity [@goldin2016; @botev2019]. [@chang2019] provided evidence through interviews with migrant workers and their family members. They said that the city offered more jobs and higher pay than did the countryside. Oftentimes, they found that labor migration presented the only way to sustain their children's education and livelihood.

The other theoretical model is the family disruption model. Children flourish under parental nurturing care. This term refers to the "stable environments that promote health and adequate nutrition, protect from threats, and provide opportunities for learning and responsive, emotionally supportive and developmentally enriching relationships" [@black, p.1]. Parental migration decreases nurturing care, thereby hindering children's developmental outcomes. The loss of nurturing care hinders children's accumulation of human capital. [@hong2019] talked to a group of left behind children in grade nine. These students expressed the utter sense of loneliness "with no advocate, no support and with no one looking out for them" (p. 13). Many of them felt indifferent to school grades, and some even longed to ditch school to pursue a free, independent life (p. 14).

@@

I obtained the data from the China Education Panel Survey (CEPS). It is a nationally representative, school-based survey featuring junior high school students. The survey administrator is the National Survey Research Center at Renmin University of China. The project's sampling adopted a multi-stage stratified probability proportional to size approach. Moreover, designed deliberately to focus on migrant children, the survey also oversampled those counties with large intranational migrant populations. The participants filled a paper-and-pencil questionnaire. The baseline survey took place in the 2013-2014 academic year with a sample size of about twenty thousand students in seventh and ninth grades. These students were nested in 438 classrooms of 112 schools in 28 urban districts or rural counties. I only kept the 9,449 observations of seventh grade students and their parents, who completed the follow-up survey in the 2014-2015 academic year.

@@

I explored two categories of adolescence development: academic and cognitive abilities. One key dependent variable, the student's cognitive skill, is measured by a fifteen-minute standardized test. The test covered three types of reasoning: verbal, visuospatial, and numerical. The raw scores of cognitive skill were then standardized. The other set of dependent variables concerned students' academic performance. This was measured by the mid-term exam scores in three subjects: Chinese, mathematics, and English. To facilitate comparison across schools and counties, I used standardized test scores instead of raw test scores. I removed the observations with missing values in any of the test score mentioned above.

The key independent variable, the parental migration status, was identified by two items in the parent survey. In both the baseline and the follow-up surveys, parents specified the family members living in the same household at the time. I identified seven types of migration arrangement: both parents absent, only father absent, only mother absent, father absent (unconditional), mother absent (unconditional), and both parents at home. I then removed the observations that reported other types of parental absence induced by divorce or death.

@@

Following Bai and colleagues' [-@bai2018; -@bai2020] approach, I adopted a difference-in-differences design to analyze the educational outcome of children newly left behind by parents vis-a-vis that of children staying together with parents. The model is: $$\Delta score_{i,s} = \alpha + \beta \cdot migr_{i,s} + \gamma \cdot FE_{s} + \lambda \cdot score_{i,s;base} + \theta \cdot X_{i,s} + \varepsilon_{i,s}$$For student $i$ in school $s$, $\Delta score_{i,s}$ is the change in standardized test score between baseline and follow-up surveys, $migr_{i,s}$ is the treatment dummy variable, $FE_{s}$ is the school-level fixed effect, $score_{i,s;base}$ is the baseline exam score, and $X_{i,s}$ is the vector of covariates capturing the characteristics of children and their households.

The control variables at the student level include: student's age, gender, ethnicity, self-rated health, hukou type, and the number of siblings. I obtained the age by subtracting the student's birth year from the year 2013, the start of the baseline academic year. I coded students' gender as male = 0 and female = 1, ethnic minority status as Han Chinese = 0 and other ethnicity = 1, hukou status as rural (agricultural) hukou = 1 and other types = 0. I obtained the number of siblings of students through two items. Students who answered yes to the question "Are you the only child of your family" had no sibling. Those who answered no to this question were instructed to proceed to the next question, "How many full or half siblings do you have" -- elder brother, younger brother, elder sister, and younger sister, respectively. In these four columns, I transformed `NA` to zero. Then I added the four columns as the number of siblings of a student.

The control variables include the internet assess at home, the self-rated number of books at home, and the years of education of the best-educated parent. These variables can approximate the level of human capital in a household. The internet assess was measured by the item "Do [sic] your family own a computer and have an access to the Internet". Those who answered "Yes, we have both" were coded as 1, and others as 0. The number of books was measured by the item "How many books do your family own? (not including textbooks or magazines)". Students chose from a range of 1 as "very few" to 5 as "a great number". The years of education of the best-educated parent were measured by two items. Students indicated their father's and mother's educational level, respectively. I transformed these two variables into years of education and retained the larger value.

I adopted `r cite_r("r-references.bib")` for all the analyses and the `fixest` package [@bergé2018] for building the two-way fixed effect models.

@@

```{r}
table_adj_unres_cog
```

```{r}
table_adj_unres_chn
```

```{r}
table_adj_unres_mat
```

```{r}
table_adj_unres_eng
```
